SI: Settlements

### 2021-09-15 analysis/paper/paper.Rmd
update paper, PCA, hclust + dendro

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

```{r packages}
devtools::load_all(".")
library(sf)
library(tidyverse)
```

@@

The analysis is divided into two 'time' scales:

@@

settlements1 <- set_spat %>%
  right_join(set_base$period1, by = "id")
settlements2 <- set_spat %>%
  right_join(set_base$period2, by = "id")
```

@@

```{r tab-set1-count}
set_base$period1 %>%
  group_by(period_label) %>%
  summarize(n = n()) %>% 
  rename('Period' = period_label) %>% 
  knitr::kable(caption = "Number of settlements across time periods.") %>% 
  kableExtra::kable_styling(full_width = FALSE)
```

```{r plot-set1-count, fig.cap = 'Settlements classified by general pottery traditions across time periods.'}
settlements1 %>%
  ggplot() +
  geom_sf(data = mask, fill = "gray80", color = NA, alpha = 0.4) +
  # geom_sf(data = admin1, linetype = 3, fill = NA, color = "gray60", size = 0.2, alpha = 0.4) +
  geom_sf(data = admin0, linetype = 1, fill = NA, color = "gray60", alpha = 0.4) +
  geom_sf(data = rivers1, color = "mediumblue", size = 0.4, alpha = 0.4) +
  geom_sf(data = rivers0, color = "mediumblue", size = 0.6, alpha = 0.4) +
  geom_sf(alpha = 0.2, aes(color = label)) +
  scale_color_brewer(palette = "Set1", name = "Tradition") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(vars(period_label), nrow = 3) +
  theme_void() +
  theme(panel.border = element_rect(color = "black", fill = NA))
```

@@

```{r tab-set2-count}
set_base$period2 %>%
  group_by(period_label) %>%
  summarize(n = n()) %>% 
  rename('Period' = period_label) %>% 
  knitr::kable(caption = "Number of settlements across time periods.") %>% 
  kableExtra::kable_styling(full_width = FALSE)
```

```{r plot-set2-count, fig.cap = 'Settlements classified by pottery groups across time periods.'}
settlements2 %>%
  ggplot() +
  geom_sf(data = mask, fill = "gray80", color = NA, alpha = 0.4) +
  # geom_sf(data = admin1, linetype = 3, fill = NA, color = "gray60", size = 0.2, alpha = 0.4) +
  geom_sf(data = admin0, linetype = 1, fill = NA, color = "gray60", alpha = 0.4) +
  geom_sf(data = rivers1, color = "mediumblue", size = 0.4, alpha = 0.4) +
  geom_sf(data = rivers0, color = "mediumblue", size = 0.6, alpha = 0.4) +
  geom_sf(alpha = 0.2, aes(color = label)) +
  scale_color_brewer(palette = "Set1", name = "Tradition") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(vars(period_label), nrow = 3) +
  theme_void() +
  theme(panel.border = element_rect(color = "black", fill = NA))
```

@@

```{r set1-kde}
kde1 <- kdestimate(settlements1,
                   mask,
                   markscol = "period",
                   lvls = names(labs_chrono$period)) %>%
  rename(period = name) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  select(id, starts_with("period"), chrono, kde)

# tibble with results
res1_long <- kde1 %>% 
  rename(settlements_kde = kde) %>% 
  select(-period_label) %>% 
  pivot_longer(col = settlements_kde, names_to = "variable")
```

```{r plot-set1-kde, fig.cap = 'Kernel density estimation for different periods.'}
kdestimate(settlements1,
           mask,
           markscol = "period",
           lvls = names(labs_chrono$period),
           points = FALSE) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  ggplot(aes(x, y, fill = kde)) +
  geom_raster() +
  scale_fill_viridis_c(name = "KDE") +
  facet_wrap(vars(period_label)) +
  coord_fixed() +
  theme_void()
```

@@

```{r set2-kde}
kde2 <- kdestimate(settlements2,
                   mask,
                   markscol = "period",
                   lvls = names(labs_chrono$period)) %>%
  rename(period = name) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  select(id, starts_with("period"), chrono, kde)

# tibble with results
res2_long <- kde2 %>% 
  rename(settlements_kde = kde) %>% 
  select(-period_label) %>% 
  pivot_longer(col = settlements_kde, names_to = "variable")
```

```{r plot-set2-kde, fig.cap = 'Kernel density estimation for different periods.'}
kdestimate(settlements2,
           mask,
           markscol = "period",
           lvls = names(labs_chrono$period),
           points = FALSE) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  ggplot(aes(x, y, fill = kde)) +
  geom_raster() +
  scale_fill_viridis_c(name = "KDE") +
  facet_wrap(vars(period_label)) +
  coord_fixed() +
  theme_void()
```

@@

```{r get_data-rm}
rm_pts <- st_read(here::here(derived_data, "rm_points.geojson"), quiet = TRUE) %>% 
  mutate(label = if_else(orig == "l", "Chipped", "Polished"))
rm_lns <- st_read(here::here(derived_data, "rm_lines.geojson"), quiet = TRUE) %>% 
  mutate(label = if_else(orig == "l", "Chipped", "Polished"))
```

@@

@@

rm_dist <- read_csv(paste0(path_temp, "rm_dist.csv"))

rm_dist_median <- rm_dist %>% 
  select(-region, -type, -shp, -min, -mean, -max) %>% 
  group_by(id, rm) %>% 
  summarise(dist_rm = mean(median)) %>% 
  pivot_longer(cols = dist_rm, names_to = "variable") %>% 
  mutate(variable = str_c(variable, rm, sep = "_")) %>% 
  select(-rm)

# update result tibbles
res1_long <- set_base$period1 %>% 
  select(id, period, chrono) %>% 
  inner_join(rm_dist_median) %>%
  bind_rows(res1_long)

res2_long <- set_base$period2 %>% 
  select(id, period, chrono) %>% 
  inner_join(rm_dist_median) %>%
  bind_rows(res2_long)
```

```{r plot-rm-dist-min, fig.cap="Distance to raw material sources."}
rm_dist %>%
  ggplot(aes(x = min, y = rm, fill = region)) +
  geom_violin() +
  scale_fill_brewer(palette = "Accent") +
  facet_wrap(~type, scales = "free_y") +
  labs(x = "min. dist. (km)", y = "raw material", fill = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r plot-rm-dist-kde-chrono1, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery traditions."}
kde1 %>% select(id, period_label, kde) %>% 
  left_join(rm_dist) %>% 
  mutate(region = str_extract(id, "^.")) %>% 
  filter(rm %in% c("AD", "ZELESICE", "PMD", "KL"),
         region == "M") %>% 
  ggplot(aes(median, kde)) +
  geom_point(alpha = 0.1) + 
  stat_smooth(method = "lm", se = FALSE) +
  facet_grid(vars(period_label), vars(rm)) +
  labs(x = "Median distance to raw material source (km)",
       y = "KDE") +
  theme_minimal() +
  theme(strip.text.y = element_text(angle = 0))
```

```{r plot-rm-dist-kde-chrono2, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery groups."}
kde2 %>% select(id, period_label, kde) %>% 
  left_join(rm_dist) %>% 
  mutate(region = str_extract(id, "^.")) %>% 
  filter(rm %in% c("AD", "ZELESICE", "PMD", "KL"),
         region == "M") %>% 
  ggplot(aes(median, kde)) +
  geom_point(alpha = 0.1) + 
  stat_smooth(method = "lm", se = FALSE) +
  facet_grid(vars(period_label), vars(rm)) +
  labs(x = "Median distance to raw material source (km)",
       y = "KDE") +
  theme_minimal() +
  theme(strip.text.y = element_text(angle = 0))
```

@@

In settlement structure studies it is often assumed that past human settlements
have some kind of a relation to the network of watercourses. Numerous problems
are faced while assessing the relationship between settlements and close
watercourse. Most importantly the man-made recent changes in the structure and
intensity of watercourses (and landscape in general) are a major factor
imapiring such analysis.

@@

```{r get-raster}
hydro_dens <- raster::raster(
  here::here(derived_data, "hydro", "density_raster.tif")) %>% 
  raster::rasterToPoints() %>%
  as_tibble() %>%
  rename(KDE = density_raster)
```

```{r plot-hydro-density, fig.cap="Kernel smoothed intensity of watercourses overlaid with sites."}
ggplot() +
  geom_raster(data = hydro_dens, aes(x, y, fill = KDE)) +
  geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
  theme_void()
```

>>>>

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = FALSE,
  comment = "#>",
  # fig.path = "../figures/",
  dpi = 300
)
```

```{r packages, cache=FALSE}
library(sf)
library(tidyverse)
library(spatstat)
library(patchwork)
devtools::load_all(".")
```

@@

The analysis is divided into two scales:

@@

settlements1 <- set_spat %>%
  right_join(set_base$period1, by = "id") %>% 
  mutate(label = labs_chrono$chrono1[chrono])
settlements2 <- set_spat %>%
  right_join(set_base$period2, by = "id") %>% 
  mutate(label = labs_chrono$chrono2[chrono])

# colors
mycol <- c("B" = "#7FC97F", "M" = "#BEAED4", "b" = "#7FC97F", "m" = "#BEAED4"
)
```

@@

```{r tab-set1-count}
set_base$period1 %>%
  table_nr_set()
```

```{r plot-set1-count, fig.cap = 'Settlements classified by general pottery traditions across time periods.'}
settlements1 %>%
  plot_nr_set()
```

@@

```{r tab-set2-count}
set_base$period2 %>%
  table_nr_set()
```

```{r plot-set2-count, fig.cap = 'Settlements classified by pottery groups across time periods.'}
settlements2 %>%
  plot_nr_set()
```

@@

```{r set1-kde}
res1_long <- settlements1 %>% 
  est_kde()
```

```{r plot-set1-kde, fig.cap = 'Kernel density estimation for different periods.'}
settlements1 %>% 
  plot_kde()
```

@@

```{r set2-kde}
res2_long <- settlements2 %>% 
  est_kde()
```

```{r plot-set2-kde, fig.cap = 'Kernel density estimation for different periods.'}
settlements2 %>% 
  plot_kde()
```

@@

```{r get_data-rm}
rm_pts <- st_read(here::here(derived_data, "rm_points.geojson"), quiet = TRUE) %>% 
  add_rm_type()

rm_lns <- st_read(here::here(derived_data, "rm_lines.geojson"), quiet = TRUE) %>% 
  add_rm_type()
```

@@

## Distances

@@

rm_dist <- read_csv(paste0(path_temp, "rm_dist.csv")) %>% 
  mutate(rm = str_to_lower(rm))

# rm_dist_mean <- rm_dist %>%
#   select(-region, -type, -shp, -min, -median, -max) %>%
#   group_by(id, rm) %>%
#   summarise(dist_rm = mean(mean)) %>%
#   pivot_longer(cols = dist_rm, names_to = "variable") %>%
#   mutate(variable = str_c(variable, rm, sep = "_")) %>%
#   select(-rm)
```

<!-- ```{r plot-rm-dist-min, fig.cap="Min. distance to raw material sources."} -->
<!-- rm_dist %>% -->
<!--   ggplot(aes(x = min, y = rm, fill = region)) + -->
<!--   geom_violin() + -->
<!--   scale_fill_brewer(palette = "Accent") + -->
<!--   facet_wrap(~type, scales = "free_y") + -->
<!--   labs(x = "min. dist. (km)", y = "raw material", fill = "Region") + -->
<!--   theme_minimal() + -->
<!--   theme(legend.position = "bottom",  -->
<!--         panel.border = element_rect(color = "black", fill = NA)) -->
<!-- ``` -->

## Settlement density against distance to raw materials

> Hypothesis: In Morava river catchment settlement position is directly associated with 
raw material source.

We compare settlement density (KDE) with distance to chosen raw materials.
There is no strong evidence that the distribution of settlements is associated with distance to 
any of the key raw material sources occurring in the region in any of the studied periods.
There are some mild negative correlations in the period 4.8 -- 4.2 k BCE for polished stone tools materials.
Strongest connection is observed between given pottery groups in period 4.8 -- 4.6 k BCE in case of ad/pmd/zelesice raw material sources.

```{r plot-rm-dist-kde-chrono1, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery traditions."}
res1_long %>% plot_dist_lm()
```

```{r tab-rm-dist-cor-kde1}
res1_long %>% tab_dist_lm()
```

```{r plot-rm-dist-kde-chrono2, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery groups."}
res2_long %>% plot_dist_lm()
```

```{r tab-rm-dist-cor-kde2}
res2_long %>% tab_dist_lm()
```

<!--
## PCA for raw material distances

Principal components analysis is used to reduce dimensionality.
First three PCs are enough to cover more than 95% variability in the data set.

```{r rm-dist-pca}
# scale_col <- function(x){
#   (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
# }
# 
# rm_dist_mean_wide <- rm_dist_mean %>%
#   mutate(variable = str_remove(variable, "dist_rm_")) %>% 
#   pivot_wider(names_from = variable, values_from = value) %>% 
#   ungroup() %>% 
#   mutate(across(where(is.numeric), scale_col))
# 
# # matrix for pca to reduce dimensions
# rm_dist_mx <- rm_dist_mean_wide %>% 
#   select(-id) %>% 
#   as.matrix() 
# row.names(rm_dist_mx) <- rm_dist_mean_wide$id
# 
# rm_pca <- prcomp(rm_dist_mx)
# summary(rm_pca)
```

```{r rm-corrplot, fig.cap="Correlation between principal components and given raw materials."}
# corrplot::corrplot(t(rm_pca$rotation[, 1:3]))
```
```{r rm-biplot, fig.cap="First thee principal components (green - East part of Bohemia, blue - Morava river catchment)."}
# pairs(rm_pca$x[, 1:3], col = unname(mycol[str_extract(rownames(rm_dist_mx), ".")]))
```

```{r update-result-rm}
# rm_dist_pca <- rm_pca$x[, 1:3] %>% 
#   as_tibble(rownames = "id") %>% 
#   rename("rm_pc1" = "PC1",
#          "rm_pc2" = "PC2",
#          "rm_pc3" = "PC3",) %>% 
#   pivot_longer(-id, "variable", "value")
# 
# # update result tibbles
# res1_long <- set_base$period1 %>% 
#   select(id, period, chrono) %>% 
#   inner_join(rm_dist_pca) %>%
#   bind_rows(res1_long)
# 
# res2_long <- set_base$period2 %>% 
#   select(id, period, chrono) %>% 
#   inner_join(rm_dist_pca) %>%
#   bind_rows(res2_long)
# 
# rm(list = ls(pattern = "^rm_dist_"))
```
-->

## RM relevant to given period and/or region

```{r rm-period}
# 0 = not used
# 0.5 = somewhat used
# 1 = used
rm_b <- readxl::read_xlsx(here::here("analysis/data/raw_data",
                                     "sidliste_vs_zdroje.xlsx"), sheet = "B")
rm_m <- readxl::read_xlsx(here::here("analysis/data/raw_data",
                                     "sidliste_vs_zdroje.xlsx"), sheet = "M")

rm_relevance <- bind_rows(b = rm_b, m = rm_m, .id = "reg") %>% 
  mutate(across(starts_with("p"), function(x) x * 0.5)) %>% 
  pivot_longer(starts_with("p"), 
               names_to = "period", values_to = "relevance") %>% 
  rename(rm = surovina)

rm_dist_rel_p1 <- set_base$period2 %>% 
  filter_rel_dist_rm()

rm_dist_rel_p2 <- set_base$period2 %>% 
  filter_rel_dist_rm()
```

### RM relations in *general* pottery traditions

#### Morava river catchment

```{r rm-dist-m-p1, fig.cap="Distance to RM sources, Morava river catchment, pottery traditions. RM sources relevant in given period in black, somewhat relevant RM sources in gray."}
rm_dist_rel_p1 %>% 
  plot_rm_reg_dist("m")
```

#### Eastern part of Bohemia

```{r rm-dist-b-p1, fig.cap="Distance to RM sources, Eastern part of Bohemia, pottery traditions. RM sources relevant in given period in black, somewhat relevant RM sources in gray."}
rm_dist_rel_p1 %>% 
  plot_rm_reg_dist("b")
```

### RM relations in *detailed* pottery groups

#### Morava river catchment

```{r rm-dist-m-p2, fig.cap="Distance to RM sources, Morava river catchment, pottery groups. RM sources relevant in given period in black, somewhat relevant RM sources in gray."}
rm_dist_rel_p2 %>% 
  plot_rm_reg_dist("m")
```

#### Eastern part of Bohemia

```{r rm-dist-b-p2, fig.cap="Distance to RM sources, Eastern part of Bohemia, pottery groups. RM sources relevant in given period in black, somewhat relevant RM sources in gray."}
rm_dist_rel_p2 %>% 
  plot_rm_reg_dist("b")
```

```{r rm-dist-rel}
rm_types <- rm_dist %>% 
  select(rm, type) %>% 
  distinct()

rm_types_tab <- rm_types %>% 
  pull(type)

names(rm_types_tab) <- rm_types %>% 
  pull(rm)

res1_long <- rm_dist_rel_p1 %>% 
  prep_dist_rel_rm() %>% 
  bind_rows(res1_long)

res2_long <- rm_dist_rel_p2 %>% 
  prep_dist_rel_rm() %>% 
  bind_rows(res2_long)
```

```{r rm-dist-rel1, fig.cap="Mean distance to relevant raw materials in pottery traditions."}
rm_dist_rel_p1 %>% 
  count_dist_rel_rm() %>% 
  plot_dist_rel_rm()
```

```{r rm-dist-rel2, fig.cap="Mean distance to relevant raw materials in pottery groups."}
rm_dist_rel_p2 %>% 
  count_dist_rel_rm() %>% 
  plot_dist_rel_rm()
```

@@

In settlement structure studies it is often assumed that past human settlements
have some kind of a relation to the network of watercourses. Numerous problems
are faced while assessing the relationship between settlements and close
watercourse. Most importantly the man-made recent changes in the structure and
intensity of watercourses (and landscape in general) are a major factor
impairing such analysis.

@@

```{r get-raster}
# pkg raster
# hydro_dens <- raster::raster(
#   here::here(derived_data, "hydro", "density_raster.tif"))
# 
# hydro_dens_at_points <- hydro_dens %>% 
#   raster::extract(set_spat) %>% 
#   as_tibble() %>% 
#   rename(hydro_kde = value) %>% 
#   bind_cols(select(st_drop_geometry(set_spat), id))

# pkg stars
hydro_dens <- stars::read_stars(
  here::here(derived_data, "hydro", "density_raster.tif")
) %>% 
  st_set_crs(value = st_crs(set_spat))

hydro_dens_at_points <- hydro_dens %>% 
  stars::st_extract(set_spat) %>% 
  st_drop_geometry() %>% 
  rename(hydro_kde = density_raster.tif) %>% 
  bind_cols(select(st_drop_geometry(set_spat), id))

# bind to results
res1_long <- set_base$period1 %>% 
  prep_hydro_dens_res() %>% 
  bind_rows(res1_long)

res2_long <- set_base$period2 %>% 
  select(-facet) %>% 
  prep_hydro_dens_res() %>% 
  bind_rows(res2_long)
```

```{r plot-hydro-density, fig.cap="Kernel smoothed intensity of watercourses overlaid with sites (raster clipped to research area extent)."}
# hydro_dens %>% 
#   raster::mask(mask) %>% 
#   raster::rasterToPoints() %>%
#   as_tibble() %>%
#   rename(KDE = density_raster) %>% 
#   ggplot() +
#   geom_raster(aes(x, y, fill = KDE)) +
#   geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
#   theme_void()

ggplot() +
  stars::geom_stars(data = st_crop(hydro_dens, mask), sf = TRUE) +
  geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
  theme_void() +
  labs(fill = "KDE")
```
<!--- ====================================================================== -->

# Terrain surrounding the settlements

We focus on slope around the settlements.

```{r dem-data}
dem_data <- read_csv(here::here(derived_data, "dem/vals_dem.csv"))

# bind to results
res1_long <- set_base$period1 %>% 
  prep_dem_res() %>% 
  bind_rows(res1_long)

res2_long <- set_base$period2 %>% 
  select(-facet) %>% 
  prep_dem_res() %>% 
  bind_rows(res2_long)
```

```{r altitude-to-res}
altitude <- set_spat %>% st_drop_geometry() %>% 
  as_tibble() %>% 
  select(id, altitude)

# bind to results
res1_long <- set_base$period1 %>% 
  prep_alt_res() %>% 
  bind_rows(res1_long)

res2_long <- set_base$period2 %>% 
  select(-facet) %>% 
  prep_alt_res() %>% 
  bind_rows(res2_long)

```

```{r dem-analysis-correlation, fig.cap='Correlation between altitude derived from DEM and altitude reading in the topographic map.'}
alt_plot <- set_spat %>% st_drop_geometry() %>% 
  as_tibble() %>% 
  select(id, altitude) %>% 
  left_join(dem_data)

alt_cor <- alt_plot %>% 
  summarise(cor = cor(altitude, dem)) %>% 
  pull(cor) %>% 
  round(2)

alt_plot %>% 
  ggplot(aes(altitude, dem)) +
  geom_point() +
  theme_minimal() +
  annotate(geom = "text", x = 175, y = 525, 
           label = paste("Cor. ", alt_cor))
```

<!--- ====================================================================== -->

# Settlement 'hierarchy'

Minimal distance to fortified settlements in given regions is explored.

```{r dist-fenced-input}
dist_fenced <- read_rds(here::here(derived_data, "fortif/distance_fortif.RDS"))

# results 1
res1_long <- dist_fenced$traditions %>% 
  prep_dist_fenced() %>% 
  bind_rows(res1_long)

# results 2
dist_fenced_p2 <- dist_fenced$groups %>% 
  prep_dist_fenced() %>% 
  bind_rows(res2_long)
```

<!--- ====================================================================== -->

# Settlement 'continuity'

Is the same location settled in the previous period?

```{r continuity-calc}
res1_long <- set_base$period1 %>%
  settlement_continuity() %>% 
  bind_rows(res1_long)

res2_long <- set_base$period2 %>%
  settlement_continuity() %>% 
  bind_rows(res2_long)
```

```{r continuity-plot1, fig.cap="Settlements continuity, pottery traditions."}
n_set_per1 <- set_base$period1 %>% 
  add_region() %>% 
  group_by(period_label, reg) %>% 
  summarise(sum = n())

res1_long %>% 
  plot_continuity(y = n_set_per1)
```

```{r continuity-plot2, fig.cap="Settlements continuity, pottery groups."}
n_set_per2 <- set_base$period2 %>% 
  add_region() %>% 
  group_by(period_label, reg) %>% 
  summarise(sum = n())

res2_long %>% 
  plot_continuity(y = n_set_per2)
```

<!--- ====================================================================== -->

# Arrangement along natural lines

Arrangement of settlements along natural 'terrain' lines or along waterways is explored.

```{r along_lines}
along_lines <- read_rds(here::here(derived_data, "lines/linear_arrangement.RDS"))

along_lines1 <- along_lines$traditions %>% 
  mutate(chrono = str_remove(chrono, ".$")) %>% 
  group_by(id, chrono, period, variable) %>% 
  summarise(value = max(value)) %>% 
  ungroup(id, chrono, variable) %>% 
  mutate(value = value / n()) %>% 
  select(-chrono) %>% 
  group_by(variable, period, id) %>% 
  summarise(value = mean(value), .groups = "drop")

along_lines2 <- along_lines$groups %>% 
  distinct(id, chrono, period, variable, value) %>% 
  group_by(period) %>% 
  mutate(value = value / n()) %>% 
  select(-chrono) %>% 
  group_by(variable, period, id) %>% 
  summarise(value = mean(value), .groups = "drop")

# results 1
res1_long <- res1_long %>% 
  bind_rows(along_lines1)

# results 2
res2_long <- res2_long %>% 
  bind_rows(along_lines2)
```

```{r clean}
rm(list = ls(pattern = "^rm"))
rm(list = ls(pattern = "^along"))
rm(list = ls(pattern = "line$"))
rm(list = ls(pattern = "violin$"))
rm(list = ls(pattern = "^n_set"))
rm(list = ls(pattern = "^hydro"))
rm(list = ls(pattern = "fenced$"))
rm(list = ls(pattern = "dem_data"))
rm(list = ls(pattern = "^alt"))
```

<!--- ====================================================================== -->

# Synthesis

```{r res-cleaning}
res1_wide <- res1_long %>% prep_wide_res()
# res1_long <- res1_wide %>% 
#   pivot_longer(-c(id, period), names_to = "variable", values_to = "value")

res2_wide <- res2_long %>% prep_wide_res()
# res2_long <- res2_wide %>% 
#   pivot_longer(-c(id, period), names_to = "variable", values_to = "value")

# # scale cols
# res1_wide <- res1_wide %>% 
#   mutate(across(where(is.numeric), scale))
# 
# res2_wide <- res2_wide %>% 
#   mutate(across(where(is.numeric), scale))
```

## Violin plots

```{r variable-labels}
labs_variables <- c(
  "settlements_kde" = "Settlement density",
  "dist_fenced" = "Min. dist. to fenced set. (km)",
  "line_terrain" = "Along terrain line",
  "line_water" = "Along water line",
  "cont" = "Settlement continuity",
  "hydro_kde" = "Watercourse density",
  # "dem" = "Altitude (DEM)", 
  "altitude" = "Altitude (map)",
  "slope" = "Slope (°)",
  "rm_dist_chipped" = "Distance to chipped RM",
  "rm_dist_polished" = "Distance to polished RM"
  # "rm_pc1" = "Distance to RM sources 1",
  # "rm_pc2" = "Distance to RM sources 2",
  # "rm_pc3" = "Distance to RM sources 3"
) %>% 
  map_chr(str_wrap, 12)
```

```{r syn-p1, fig.height=12}
res1_violin <- res1_long %>% plot_violin()
res1_violin
```

```{r syn-p2, fig.height=12}
res2_violin <- res2_long %>% plot_violin()
res2_violin
```

```{r res-violin-out}
ggsave(here::here("analysis/figures", "res1_violin.png"), res1_violin, 
       width = 9, height = 12)
ggsave(here::here("analysis/figures", "res2_violin.png"), res2_violin, 
       width = 9, height = 12)
```

## Line plots

```{r syn-p1-sum, fig.cap="Summaries for different variables across studied periods. Solid line - mean value, wide transparent line - median value.", fig.height=12}
res1_line <- res1_long %>% plot_lines()
res1_line
```

```{r syn-p2-sum, fig.cap="Summaries for different variables across studied periods. Solid line - mean value, wide transparent line - median value.", fig.height=12}
res2_line <- res2_long %>% plot_lines()
res2_line
```

```{r res-line-out}
ggsave(here::here("analysis/figures", "res1_line.png"), res1_line, 
       width = 7, height = 7)
ggsave(here::here("analysis/figures", "res2_line.png"), res2_line, 
       width = 7, height = 7)
```

<!-- ## PCA -->

<!-- ### Pottery traditions -->

<!-- ```{r pca1-all} -->
<!-- # fit PCA -->
<!-- pca1_all_fit <- res1_wide %>%  -->
<!--   # select(-cont) %>%  -->
<!--   select(where(is.numeric)) %>%  -->
<!--   scale(center = TRUE, scale = TRUE) %>%  -->
<!--   prcomp() -->

<!-- # augmented fitted values -->
<!-- pca1_all_aug <- pca1_all_fit %>%  -->
<!--   broom::augment(res1_wide) %>%  -->
<!--   select(-".rownames") %>%  -->
<!--   rename_with(\(x) str_remove(x, ".fitted")) %>%  -->
<!--   settlements::add_region() %>%  -->
<!--   settlements::add_period_label() %>%  -->
<!--   settlements::add_neo() -->

<!-- # rotation matrix -->
<!-- pca1_all_rot <- pca1_all_fit %>% -->
<!--   broom::tidy(matrix = "rotation") %>% -->
<!--   mutate(PC = str_c("PC", PC)) %>% -->
<!--   pivot_wider(names_from = PC) -->

<!-- # var explained -->
<!-- pca1_all_var_exp <- pca1_all_fit %>%  -->
<!--   broom::augment(res1_wide) %>% -->
<!--   summarise(across(contains("PC"), var)) %>%  -->
<!--   pivot_longer(cols = everything(), names_to = "pc", values_to = "variance") %>%  -->
<!--   mutate(var_exp = variance/sum(variance), -->
<!--          cum_var_exp = cumsum(var_exp), -->
<!--          pc = str_replace(pc, ".fitted", "")) %>%  -->
<!--   select(pc, var_exp, cum_var_exp) -->

<!-- # PC cum var explained over 0.9 -->
<!-- pca1_all_var_exp_over <- pca1_all_var_exp %>%  -->
<!--   mutate(n = !(cum_var_exp >= 0.9)) %>%  -->
<!--   pull(n) %>%  -->
<!--   sum() + 1 -->
<!-- ``` -->

<!-- ```{r pca1-all-var-exp-tab} -->
<!-- knitr::kable(pca1_all_var_exp, col.names = c("Comp.", -->
<!--                                              "Prop. of variance",  -->
<!--                                              "Cumulative prop.")) -->
<!-- ``` -->

<!-- ```{r pca1-all-loadings-plot} -->
<!-- # PCA loadings -->
<!-- pca1_all_loadings <- pca1_all_fit %>%  -->
<!--   broom::tidy(matrix = "rotation") %>%  -->
<!--   mutate(component = str_c("PC", PC), -->
<!--          component = fct_reorder(component, PC)) %>%  -->
<!--   filter(PC <= pca1_all_var_exp_over) %>%  -->
<!--   group_by(PC) %>%  -->
<!--   top_n(8, abs(value)) %>%  -->
<!--   ungroup() %>%  -->
<!--   mutate(cor = if_else(value > 0, "positive", "negative"), -->
<!--          column = tidytext::reorder_within(column, abs(value), component)) -->

<!-- pca1_all_loadings %>%  -->
<!--   ggplot(aes(abs(value), column, fill = cor)) + -->
<!--   geom_col(color = "black") + -->
<!--   scale_fill_manual(values = c("white", "gray80")) + -->
<!--   facet_wrap(vars(component), scales = "free_y") + -->
<!--   tidytext::scale_y_reordered() + -->
<!--   theme_bw() + -->
<!--   labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") -->
<!-- ``` -->

<!-- ```{r plot-pca1-results} -->
<!-- plot_pca1_full <- function(x = "PC1", y = "PC2") { -->

<!--   pcs <- c(x, y) -->

<!--   pca1_all_plot <- pca1_all_aug %>%  -->
<!--     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) + -->
<!--     geom_point(data = select(pca1_all_aug, -period_label),  -->
<!--                alpha = 0.05, size = 0.4) + -->
<!--     geom_point(aes(color = reg),  -->
<!--                alpha = 0.8, size = 0.5, show.legend = FALSE) + -->
<!--     scale_color_manual(values = mycol[c(1, 2)]) + -->
<!--     facet_grid(vars(period_label), vars(reg)) + -->
<!--     theme_minimal() -->

<!--   pca1_all_rot <-  pca1_all_rot %>% -->
<!--     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) + -->
<!--     geom_segment(xend = 0, yend = 0, alpha = 0.4) + -->
<!--     geom_text( -->
<!--       aes(label = column), -->
<!--       hjust = 1, nudge_x = -0.02, -->
<!--       color = "#904C2F", -->
<!--       size = 2.6 -->
<!--     ) + -->
<!--     theme_minimal() -->

<!--   pca1_all_load <- pca1_all_loadings %>%  -->
<!--     filter(component %in% pcs) %>%  -->
<!--     ggplot(aes(abs(value), column, fill = cor)) + -->
<!--     geom_col(color = "black", show.legend = FALSE) + -->
<!--     scale_fill_manual(values = c("white", "gray80")) + -->
<!--     facet_wrap(vars(component), scales = "free_y") + -->
<!--     tidytext::scale_y_reordered() + -->
<!--     labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") + -->
<!--     theme_minimal() -->

<!--   pca1_all_plot / (pca1_all_rot | pca1_all_load) +  -->
<!--     plot_layout(heights = c(6, 1)) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r pca1-all-plots-pc1-pc2} -->
<!-- plot_pca1_full() -->
<!-- ggsave(here::here("analysis/figures/pca1_full_pc12.png"), height = 14) -->
<!-- ``` -->

<!-- ```{r pca1-all-plots-pc3-pc4} -->
<!-- plot_pca1_full("PC3", "PC4") -->
<!-- ggsave(here::here("analysis/figures/pca1_full_pc34.png"), height = 14) -->
<!-- ``` -->

<!-- ```{r pca1-all-plots-pc5-pc6} -->
<!-- plot_pca1_full("PC5", "PC6") -->
<!-- ggsave(here::here("analysis/figures/pca1_full_pc56.png"), height = 14) -->
<!-- ``` -->

<!-- ```{r pca1-all-plots-pc7-pc8} -->
<!-- plot_pca1_full("PC7", "PC8") -->
<!-- ggsave(here::here("analysis/figures/pca1_full_pc78.png"), height = 14) -->
<!-- ``` -->

<!-- ### Pottery groups -->

<!-- ```{r pca2-all} -->
<!-- # fit PCA -->
<!-- pca2_all_fit <- res2_wide %>%  -->
<!--   # select(-cont) %>%  -->
<!--   select(where(is.numeric)) %>%  -->
<!--   scale(center = TRUE, scale = TRUE) %>%  -->
<!--   prcomp() -->

<!-- # augmented fitted values -->
<!-- pca2_all_aug <- pca2_all_fit %>%  -->
<!--   broom::augment(res2_wide) %>%  -->
<!--   select(-".rownames") %>%  -->
<!--   rename_with(\(x) str_remove(x, ".fitted")) %>%  -->
<!--   settlements::add_region() %>%  -->
<!--   settlements::add_period_label() %>%  -->
<!--   settlements::add_neo() -->

<!-- # rotation matrix -->
<!-- pca2_all_rot <- pca2_all_fit %>% -->
<!--   broom::tidy(matrix = "rotation") %>% -->
<!--   mutate(PC = str_c("PC", PC)) %>% -->
<!--   pivot_wider(names_from = PC) -->
